\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-30 tex/paper.tex
save

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/multi.pdf}
    \caption{Multidimensional continuous data.
             ({\it Left}) Likelihood $q(\theta|\psi)$ at the beginning of the procedure, for a proposal distribution initialized with zero mean and unit variance.
             ({\it Right}) Likelihood $q(\theta|\psi)$ after adverarial variational optimization. The proposal density correctly concentrates around $\theta^* = (1, -1)$.
             }\label{fig:multi}
\end{figure}

@@

As a more challenging example, we now consider a (simplified) simulator from
particle physics for electron--positron collisions resulting in muon--antimuon
pairs ($e^+e^- \rightarrow \mu^+\mu^-$). The simulator approximates the
distribution of observed measurements $\mathbf{x} = \cos(A) \in [-1,1]$, where $A$ is the
polar angle of the outgoing muon with respect  to the originally incoming
electron. This random variable is approximately distributed as
\begin{equation}
    p(\mathbf{x}|E_\text{beam}, G_f) = \frac{1}{Z} \left[ (1 + \mathbf{x}^2) + c(E_\text{beam}, G_f) \mathbf{x} \right]
\end{equation}
where $Z$ is a known normalization constant and $c$ is an asymmetry coefficient
function. Due to the linear term in the expression, the density $p(\mathbf{x} |
E_\text{beam}, G_f)$ exhibits a so-called {\it forward-backward} asymmetry.  Its
size depends on the values of the parameters $E_\text{beam}$ (the beam energy)
and $G_f$ (the Fermi constant) through the coefficient function $c$. Given the
known analytical form of the probability density function, a simulator for
$\mathbf{x}$ can be implemented with rejection sampling~\citep{von195113}. As an
arbitrary number of random drawings may be required before passing the
acceptance threshold, this illustrates how the latent variable
$\mathbf{z}$ does not necessarily need to be fixed-sized vector, and how its
density  $p(\mathbf{z}|\theta)$ may depend on $\theta$ in some complicated way
(Section~\ref{sec:problem}).

@@

The top left plot in Figure~\ref{fig:poisson} illustates the resulting proposal
distributions $q(\theta|\psi)$ for $\gamma=0$ and $\gamma=5$ after adversarial
variational optimization. We see that the distributions both arrive in the
neighborhood of $\theta^*$, with a density highly more concentrated for
$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
distance with $\theta^*$, both models closely match the observed data, as shown
in the top right plot of  Figure~\ref{fig:poisson}, with again
slighlty better results for the entropy penalized model. This suggests either a
relatively flat landscape around $\theta^*$ or that the observed data can in
this case also be reproduced with a smeared variation $p(\mathbf{x}|\psi)$ of
the simulator. Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that
for both $\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly
fall towards $0$, which indicates  convergence towards a distribution that the
critic cannot distinguish from the true distribution.

@@

In this note, we develop a likelihood-free inference algorithm for fitting a
forward non-differentiable simulator to continuous or discrete observed data.
The algorithm combines adversarial training  with variational optimization to
minimize variational upper bounds  on the otherwise non-differentiable
adversarial objectives.
Effectively, the procedure results in learning an abitrarily tight proposal
distribution over the simulator parameters, such that corresponding the marginal
distribution of the generated data matches the observations. Preliminary results
with non-differentiable discrete or continuous simulators validate the proposed method.

>>>>

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/multi.pdf}
    \caption{Multidimensional continuous data.
             ({\it Left}) Likelihood $q(\theta|\psi)$ at the beginning of the procedure, for a proposal distribution initialized with zero mean and unit variance.
             ({\it Right}) Likelihood $q(\theta|\psi)$ after adversarial variational optimization. The proposal density correctly concentrates around $\theta^* = (1, -1)$.
             }\label{fig:multi}
\end{figure}

@@

As a more challenging example, we now consider a (simplified) simulator from
particle physics for electron--positron collisions resulting in muon--antimuon
pairs ($e^+e^- \rightarrow \mu^+\mu^-$). The simulator approximates the
distribution of observed measurements $\mathbf{x} = \cos(A) \in [-1,1]$, where $A$ is the
polar angle of the outgoing muon with respect  to the originally incoming
electron. This random variable is approximately distributed as
\begin{equation}
    p(\mathbf{x}|E_\text{beam}, G_f) = \frac{1}{Z} \left[ (1 + \mathbf{x}^2) + c(E_\text{beam}, G_f) \mathbf{x} \right]
\end{equation}
where $Z$ is a known normalization constant and $c$ is an asymmetry coefficient
function. Due to the linear term in the expression, the density $p(\mathbf{x} |
E_\text{beam}, G_f)$ exhibits a so-called {\it forward-backward} asymmetry.  Its
size depends on the values of the parameters $E_\text{beam}$ (the beam energy)
and $G_f$ (the Fermi constant) through the coefficient function $c$. Given the
known analytic form of the probability density function, a simulator for
$\mathbf{x}$ can be implemented with rejection sampling~\citep{von195113}. As an
arbitrary number of random drawings may be required before passing the
acceptance threshold, this illustrates how the latent variable
$\mathbf{z}$ does not necessarily need to be fixed-sized vector, and how its
density  $p(\mathbf{z}|\theta)$ may depend on $\theta$ in some complicated way
(Section~\ref{sec:problem}).

@@

The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting proposal
distributions $q(\theta|\psi)$ for $\gamma=0$ and $\gamma=5$ after adversarial
variational optimization. We see that the distributions both arrive in the
neighborhood of $\theta^*$, with a density highly more concentrated for
$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
distance with $\theta^*$, both models closely match the observed data, as shown
in the top right plot of  Figure~\ref{fig:weinberg}, with again
slightly better results for the entropy penalized model. This suggests either a
relatively flat landscape around $\theta^*$ or that the observed data can in
this case also be reproduced with a smeared variation $p(\mathbf{x}|\psi)$ of
the simulator. Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that
for both $\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly
fall towards $0$, which indicates  convergence towards a distribution that the
critic cannot distinguish from the true distribution.

@@

In this note, we develop a likelihood-free inference algorithm for fitting a
forward non-differentiable simulator to continuous or discrete observed data.
The algorithm combines adversarial training  with variational optimization to
minimize variational upper bounds  on the otherwise non-differentiable
adversarial objectives.
Effectively, the procedure results in learning an arbitrarily tight proposal
distribution over the simulator parameters, such that corresponding the marginal
distribution of the generated data matches the observations. Preliminary results
with non-differentiable discrete or continuous simulators validate the proposed method.
